\section{Introduction to Barendregt's Lambda Cube}
\subauthor{Silvia Ghilezan}{University of Novi Sad}
\plquote{``The looks of the pizza are the syntax and the flavor is the semantics''}
\begin{abstract}
	The Lambda Cube is a framework introduced by Henk Barendregt to systematically investigate eight different systems of lambda calculi with types. The basic system is the lambda calculus with simple types and all other systems are obtained as its three-dimensional extensions with: polymorphic types, type operators and dependent types.
\end{abstract}

\subsection{Overview}
The ``lambda cube'' consists of 3 possible additions to the basic lambda type system $\lambda\to$:
\begin{itemize}
	\item $\lambda2$ - Polymorphic types
	\item $\lambda$\underbar{$\omega$} (read: ``weak $\lambda\omega$'') - Type constructors
	\item $\lambda P$ - Dependent types
\end{itemize}

\subsection{Background}
Leibiniz sought to create a ``Characteristica universalis'' - a universal language for describing all mathematical knowledge.
In 1928, Hilbert and Ackerman published their paper on the Entscheidungsproblem or ``decision problem'': Given all axioms
of math, is there an algorithm that can tell if a proposition is deducable from the axioms?

In the 1930s, negative answers were given to the Entscheidungsproblem. G\"odel gave the first negative result in 1931
with his ``incompleteness theorems''. Later, Church gave another negative answer: it is undecidable to check if
two \lc terms are indecidable. Around that time, Turing proved that the halting problem for Turing
machines is also undecidable.

Multiple authors worked on the problem at roughly the same times and came up with similiar solutions.

Church was motivated by the formalization of mathematics. He came up with a successful model for computable
functions: the \lc. Later, types were added to create the simply typed \lc: $\lambda\to$.

Curry was motivated by the elimination of variables in logic. He came up with a successful model for computable
functions: combinatory logic. Later, types were added to create combinatory logic with types.

Turing was motivated by formalization of algorithms and reasoning about them. He came up with a successful model for computable
functions: the Turing machine.

In the mid 1930s, the equivalence of these models (as well as the model of recursive functions) was proven by Curry, Kleene and
Turing.

\subsection{Lambda Calculus Review}
Syntax:
\[
M ::= x \mid c \mid (M M) \mid (\lambda x. M)
\]
Where $x\in V$, a countable set of variables, and $c\in C$, a countable set of constants. The \lc is ``pure'' if $C = \emptyset$.

\paragraph{Conventions for associativity}
\begin{itemize}
	\item $M_1 M_2 M_3$ stands for $((M_1 M_2) M-3)$ - application is left-associative.
	\item $\lambda xy. M$ is shorthand for $(\lambda x. (\lambda y. (M)))$ - abstraction is right-associative.
	\item $\lambda x. M N$ is shorthand for $(\lambda x. (M N))$ - application takes priority over absraction.
\end{itemize}

\paragraph{Combinators}
\begin{itemize}
	\item \textbf{I} $\equiv \lambda x. x$ - combinator \textbf{I}
	\item \textbf{K} $\equiv \lambda xy. x$ - combinator \textbf{K}
	\item \textbf{S} $\equiv \lambda xyz. xz (yz)$ - combinator \textbf{S}
	\item \textbf{$\Delta$} $\equiv \lambda x. xx$ - self application
	\item \textbf{Y} $\equiv \lambda f. (\lambda x. f (x x)) (\lambda x. f (x x))$ - fixed point combinator 
	\footnote{In \lc, every function has exactly one fixed point.}
	\item \textbf{$\Omega$} $\equiv \Delta \Delta \equiv (\lambda x. x x)(\lambda x. x x)$ - higher-order function
\end{itemize}

\paragraph{Free and bound variables}

The set of free variables of the \lc term $M$ is defined inductively.
\begin{itemize}
	\item $FV(x) = {x}$
	\item $FV(MN) = FV(M)\cup FV(N)$
	\item $FV(\lambda x. M) = FV(M) \setminus {x}$
\end{itemize}
A variable in $M$ is bound if it now free. $M$ is a closed \lc term (or ``combinator'') if $FV(M) = \emptyset$. $\Lambda^\circ$ denotes 
the set of closed \lc terms.

\paragraph{Reduction rules - operational semantics}
\begin{itemize}
	\item $\alpha$-reduction: $\lambda x. M \reduces{\alpha} \lambda y. M[x := y]$ for all $y\notin FV)M)$. $\alpha$-reduction expresses
	the principle that the name of the bound variable is irrelevant. 
	$\xreduces{\alpha}$ is an equivalence relation and is denoted $\equiv_\alpha$.
	In this context is is also called ``$\alpha$-conversion''.
	\item $\beta$-reduction: $(\lambda x. M) N \reduces{\beta} M[x := N]$. $\beta$-reduction is a formalization of function evaluation.
	$\beta$-conversion which is the symmetric closure of $\xreduces{\beta}$ with $\alpha$-reduction is an equivalence relation  and
	is denoted $\equiv_\beta$.
	\emph{Barendregt's variable convention}: If a term contains a free variable which would become bound after $\beta$-reduction,
	that variable should be renamed. Renaming could be done also by using \emph{De Bruijn name-free notation}.
	For example, $(\lambda xy. xy)(yz)$ should have the bound $y$ variable renamed using $\alpha$-reduction.
	\item $\eta$-reduction: $\lambda x. (M x) \reduces{\eta} M$ for all $x\notin FV(M)$. $\eta$-reduction
	expresses \emph{extensionality}: two functions are equal if and only if the results of their application on any parameter are equal.
\end{itemize} 

\paragraph{Properties}
Confluence - the order of reductions does not matter.

\textbf{Theorem (Church-Rosser)}: For $\tau\in{\alpha, \beta, \eta}$ we have that if $M \xreduces{\tau} N$ and $M \xreduces{P}$ then
there exists $S$ such that $N \xreduces{\tau} S$ and $P \xreduces{\tau} S$.

The proof is deep and involved.

Corollaries:
\begin{itemize}
\item If $M \xreduces{\tau} N$ and $M \xreduces{P}$ then $N \equiv_\tau P$.
\item The order of reductions does not matter and always leads to the same result.
\item Reductions can be executed in parallel.
\end{itemize}

Normal forms: \lc terms that cannot be reduced any further. $P\in\Lambda$ is \textbf{normalizing} if it has a normal form. 
It is \textbf{strongly-normalizing} if all reductions of $P$ are finite. 
We define $\xreduces \equiv \xreduces{\alpha} \cup \xreduces{\beta}$.

\textbf{Theorem (normalization)}: Each \lc term has at most one normal form, up to $\alpha$-conversion.

\paragraph{Logic, conditionals and pairs}

Propositional logic in \lc:
\[
\top := \lambda xy. x,~~~~\bottom := \lambda xy.y,~~~~\neg := \lambda x. x \top \bottom
\wedge := \lambda xy. xy\bottom,~~~~\vee := \lambda xy. x\top y
\]

Pairs:
\[
pair x y := \lambda p. p x y,~~~~fst p := p \top,~~~~snd p := p \bottom
\]

Arithmetic - Church numerals:
\[
0 := \lambda fx. x
1 := \lambda fx. fx
n := \lambda fx. f^n x
add := \lambda xyz. xp (ypq)
mult := \lambda xyz. x (yz)
succ := \lambda xyz. xy(yz)
exp := \lambda xy. yz
iszero := \lambda n. n (\top \bottom) \top
\]
\textbf{Exercise}: $add n m \equiv_\beta n + m$, $mult n m \equiv_\beta n + m$

\subsection{Simple Functional Types}
The untyped \lc has its disadvantages: there exist terms without normal forms, and it is possible
to apply functions in a meaningless way, such as $+ +$. 

\emph{Types} are syntactical objects that can be assigned to $\lambda$ terms. There are two main
paradigms for simply-typed \lc:
\begin{itemize}
	\item \'a la Church - explicit type assignment (aka: typed \lc)
	\item \'a la Curry - implicit type assignment (aka: \lc with types)
\end{itemize}

\paragraph{Types}
\[
\sigma ::= \alpha | (\sigma \to \sigma)
\]

Where $\alpha \in \mathrm{TVar}$ is a countable set of type variable. In addition, we agree
that $\sigma_1 \to \sigma_2 \to \sigma_3$ means $(\sigma_1 \to (\sigma_2 \to \sigma_3))$

We have \emph{type assignments}: $M : \sigma$, where $M$ is a \lc term and $\sigma$ is a type. If $M$ is a variable,
we also call this a declaration. We have a type environment: $\Gamma$ which is a set of type assignments.

Axiom:
\[
\frac{}{\Gamma, x : \sigma \vdash x : \sigma}
\]

Rules:
\[
\frac{\Gamma \vdash M : \sigma \to \tau~~~~\Gamma \vdash N : \sigma}{\Gamma \vdash M N : \tau}
\]
For \'a la Church:
\[
\frac{\Gamma, x: \sigma \vdash M : \tau}{\Gamma : \lambda x: \sigma. M : \sigma \to \tau}
\]
For \'a la Curry:
\[
\frac{\Gamma, x: \sigma \vdash M : \tau}{\Gamma : \lambda x. M : \sigma \to \tau}
\]

Note: simply typed \lc is not Turing complete - it does not admit programs that do not terminate.

Uniqueness of types: If $\Gamma \vdash M : \sigma$ and $\Gamma \vdash M : \tau$ then $\sigma = \tau$.

The Church-Rosser property holds in the simply typed \lc as well.

Subject reduction - type preservation under reduction: If $M \reduces{} P$ and $M : \sigma$, then $P : \sigma$. 
This means that evaluation of terms does not cause types to change. In other words, types are \emph{sound}
under reduction.

\textbf{Theorem (Strong Normalization, Tait 1967):} If $M : \sigma$, then $M$ is strongly normalizing.

This theorem was proved using the \emph{reducibility method} which later became \emph{reducability candidates}
and \emph{logical relations} \Cref{sec:logicalrels}.

\subsection{Expressiveness of Simply-Typed Lambda Calculus}
Self application is not typable. Numerals are typable.

\textbf{Definition (Extended Polynomials)}: The smallest class of functions including $0, 1$, projections,
addition, multiplication and $\mathrm{ifzero}(n, m, p) := \mathrm{if} n = 0 \mathrm{then} m \mathrm{else} p$.

\textbf{Theorem}: $M$ is typable in $\lambda \to$ iff $M$ is an extended polynomial.

\paragraph{$\lambda\to$ and logic}
Logic can come in three forms:
\begin{itemize}
\item Axiomatic (Hilbert-style) which relies on axioms.
\item Natural deduction which has one axiom and rules of elimination and introduction.
\item Sequent calculus which relies on rules of elimination and introduction.
\end{itemize}

For \lc, the most natural way to give logic is the natural deduction style. 

These are intuitionistic logic rules (minimal):

Axiom:
\[
\frac{}{\Gamma, \sigma \vdash \sigma}
\]

Rules:
\[
\begin{array}{cc}
(\to_\mathrm{elim})&\frac{\Gamma \vdash \sigma \to \tau~~~~\Gamma \vdash \sigma}{\Gamma \vdash \tau} \\
(\to_\mathrm{intr})&\frac{\Gamma, \sigma \vdash \tau}{\Gamma \vdash \sigma \to \tau}
\end{array}
\]

The elimination rule is also called \emph{Modus Ponens} in Aristotelian terms. For example, let's prove $\vdash \alpha \to \alpha$.

We can use the axiom with an empty $\Gamma$, and deduce $\alpha \vdash \alpha$. Next, we apply the introduction
rule: $\alpha \vdash \alpha$ and so $\vdash \alpha \to \alpha$.

These rules correspond to rules phrased in the language of \lc and types:

Axiom:
\[
\frac{}{\Gamma, x: \sigma \vdash x: \sigma}
\]

Rules:
\[
\begin{array}{cc}
(\to_\mathrm{elim})&\frac{\Gamma \vdash x: \sigma \to \tau~~~~\Gamma \vdash y: \sigma}{\Gamma \vdash (x y): \tau} \\
(\to_\mathrm{intr})&\frac{\Gamma, x: \sigma \vdash y: \tau}{\Gamma \vdash (\lambda x. y): \sigma \to \tau}
\end{array}
\]

\paragraph{Curry-Howard Correspondence}
The Curry-Howard Correspondence is the claim that $\vdash \sigma \leftrightarrow \vdash M : \sigma$. In
words, a formula is provable in minimal intuitionistic logic if and only if it is inhabited in $\lambda\to$.

In the CHC, formulae correspond to types. Proofs are terms or programs. Proof normalization is term reduction.

We have three fundemental problems:
\begin{itemize}
\item Type checking: Given $M$ and $\sigma$, is $M : \sigma$?
\item Type inference: Given $M$, find $\sigma$ such that $M : \sigma$, or declare that one does not exist.
\item Type inhabition: Given $\sigma$, find $M$ such that $M : \sigma$, or declare that one does not exist.
\end{itemize}

In simply-typed \lc, all three problems are decidable.

ST \lc has some shortcomings: it has no self-application or recursion. As a result, it is not Turing complete.

\subsection{$\lambda2$ polymorphic types}
As we saw, the expressiveness of $\lambda\to$ is quite limited. One problem is that each function has exactly
one type and functions cannot be reused across types. The solution is to define functions that treat types
polymorphically and add $\forall \alpha. \sigma$ terms that quantify over types. For example, the type
$\forall \alpha. (\alpha \to \alpha$ is the type of all functions accepting a parameter of some type
and returning a value of the same type.

In the 1970s, Girard called this System $\mathcal{F}$ and Reynolds called it \textsc{Forsythe}.
These systems add $\lambda \alpha. \mathrm{Expression}$ and $(\mathrm{Expression} \mathrm{Type})$ to the set
of terms, and $\forall \alpha. \mathrm{Type}$ to the set of types.

There are now two types of application: it is possible to apply two terms (regular application)
or a type to a term (instantiation). There are also two kinds of $\beta$-reductions.

We also extend our typing rules:
\[
\begin{array}{c}
\frac{\Gamma \vdash M : \sigma}{\Gamma \vdash \lambda \alpha. M : \forall \alpha. \sigma} \\

\frac{\Gamma \vdash M : \forall \alpha. \sigma}{\Gamma \vdash M \tau : \sigma[\tau/\alpha]}
\end{array}
\]

Now we can type $\lambda x. x x$: it has type $(\forall \alpha. \alpha) \to \tau$.

$\lambda2$ has type uniqueness, Church-Rosser, subject reduction and strong normalization as well.

G\"odel's first incompleteness theorem from 1931 states that there exist true statements that cannot be proven.
In 1970, it was proven that strong normalization is expressible in $\lambda2$, but not provable in $\lambda2$.
It is provable in slightly stronger systems, making it a natural example of a true statement that cannot be proven.

\paragraph{Expressiveness of $\lambda2$}
We define the type of natural numbers: $\forall \alpha. \alpha \to (\alpha \to \alpha) \to \alpha$.

We can also define iteration: If $c : \sigma$ and $g: \sigma \to \sigma$, we can define $\mathrm{It} c g : \mathrm{Nat} \to \sigma$
as $\lambda n: \mathrm{Nat}. n \sigma c g$. The meaning of $\mathrm{It} c g n$ would be evaluating $g$ $n$ times beginning with $c$.

\textbf{Theorem}: $\lambda2$ types exactly all primitive recursive functions. For example, Ackerman's function is
recursive but not PR and therefore not typable in $\lambda2$.

\textbf{Theorem (Girard, Reynolds, Curry-Howard)}: A second-order logic formula holds if it is inhabited in $\lambda 2$.

It is possible to define the other connectives:
\[
\begin{array}{c}
\bottom := \forall \alpha. \alpha \\
\sigma \wedge \tau := \forall \alpha. (\sigma \to \tau \to \alpha) \to \alpha \\
\sigma \vee \tau := \forall \alpha. (\sigma \to \alpha) \to (\tau \to \alpha) \to \alpha \\
\exists \alpha. \sigma := \forall \beta. (\forall \alpha. \sigma \to \beta) \to \beta
\end{array}
\]

Constructive second order propositional logic - Peirce's law: 
$\not\vdash M : \forall \alpha. \forall \beta. ((\alpha \to \beta) \to \alpha) \to \alpha$

Peirce's law is uninhabited. It is valid in classical logic but not in intuitionistic logic.

Unfortunately, none of the three problems presented earlier is decidable for $\lambda 2$. A restriction of $\lambda 2$ has
an easy type inference algorithm: Hindley-Milner type inference.

\subsection{Dependency and \wlo}
Terms and types are mutually dependent.
\begin{itemize}
\item Terms depending on terms: $\lambda \to$
\item Terms depending on types: $\lambda 2$
\item Types depending on types: $\lambda$ \underbar{$\omega$}
\item Types depending on types: $\lambda P$
\end{itemize}


The simply typed \lc has terms depending on terms: these are just functions. We have function abstraction and function applications.
In $\lambda 2$ we also have \emph{type} abstraction: $\lambda \alpha : \star. P \alpha$ and \emph{type} application: $P\sigma$. 
Instead of saying that $\tau$ is a type, we will write $\tau : \star$.

Next, we will introduce \wlo, which has types depending on types. We define a set of ``meta-types''
called \emph{kinds}: $\mathcal{K} = {\star, \star \to \star, ...}$. We denote ``$k$ is a kind'' as $k : \square$.
Now, we can have types depending on types:  $\lambda \tau: \star. \tau \to \tau$. This expression has kind $\star \to (\star \to \star)$.
In \wlo, types and terms are kept separate. If $\sigma : *, f : \sigma$, then we say that $f$ is a \emph{constructor} for $\sigma$.

In \wlo, we introduce \emph{pseudo-expressions} as a unified framework for terms and types:
\[
\mathcal{T} ::= \mathcal{V} | \mathcal{C} | \mathcal{TT} | \lambda \mathcal{V} : \mathcal{T. T} | \mathcal{T \to T}
\]

Where ${\star, \square} = \mathcal{S} \subseteq \mathcal{C}$ is the set of \emph{sorts}. We will let $s\in {\star, \square}$ in
perpetuity.
We can now make determinations of the form $M : A$ where $M, A \in \mathcal{T}$.

The typing rules:
\begin{equation}
\begin{array}{r@{\hspace{0.4em}}c@{\hspace{0.4em}}l}
\text{(ax/sort)}&\vdash \star : \square& \\
\text{(weak)}&\frac{\Gamma \vdash M : B~~~~\Gamma \vdash A : s}{\Gamma, x: A \vdash M : B}& \text{if } x \notin \Gamma \\
\lambda&\frac{\Gamma, x. A \vdash M : B~~~~\Gamma \vdash A \to B : s}{\Gamma \vdash \lambda x: A. M : A \to B}& \\
\text{(app)}&\frac{\Gamma \vdash M : A \to B~~~~\Gamma \vdash N : A}{\Gamma \vdash MN : B}&\\
\text{(conv}_\beta\text{)}&\frac{\Gamma \vdash M : A~~~~\Gamma \vdash B : s}{\Gamma \vdash M : B}&A \equiv_\beta B \\
\text{(type/kind)}&\frac{\Gamma \vdash A : s~~~~\Gamma \vdash B : s}{\Gamma \vdash A \to B : s}&\\
\text{(var)}&\frac{\Gamma \vdash A : s}{\Gamma, x: A \vdash x : A}&\text{if } x \notin \Gamma
\end{array}
\end{equation}


Specifically, the (type/kind) rule provides that no expression may mix types and kinds, and therefore terms and types.
For example:
\[
\not\vdash \lambda \alpha: \star. \lambda \beta: \star. \lambda x : (\alpha \to \beta). x: \star \to \star \to ((\alpha \to \beta) \to (\alpha \to \beta))
\]

\wlo has the typical properties we know from other systems: uniqueness of types, Church-Rosser, subject reduction and strong normalization.
Its expressive power is exactly the same as ST \lc: Only extended polynomials can be typed.

\subsection{Types depending on terms}
We now extend the notion of ``kinds'' to include kinds such as $A \to k$ where $A : \star$ is a type and $k : \square$ is a kind.
This will allow us to have types depending on terms:
\[
f : A \to \star~~~~a: A \Rightarrow f a : \star
\]

In $\lambda P$, pseudo-expressions are defined as:
\[
\mathcal{T} ::= \mathcal{V} | \mathcal{C} | \mathcal{TT} | \lambda \mathcal{V} : \mathcal{T. T} | \Pi \mathcal{V: T. T}
\]

As before, we have ${\star, \square} = \mathcal{S} \subseteq \mathcal{C}$. The typing rules for $\lambda P$ are given here:
\begin{equation}
\begin{array}{r@{\hspace{0.4em}}c@{\hspace{0.4em}}l}
\text{(ax/sort)}&\vdash \star : \square& \\
\text{(weak)}&\frac{\Gamma \vdash M : B~~~~\Gamma \vdash A : s}{\Gamma, x: A \vdash M : B}& \text{if } x \notin \Gamma \\
\lambda&\frac{\Gamma, x. A \vdash M : B~~~~\Gamma \vdash A \to B : s}{\Gamma \vdash \lambda x: A. M : \Pi A. B}& \\
\text{(app)}&\frac{\Gamma \vdash M : \Pi x: A. B~~~~\Gamma \vdash N : A}{\Gamma \vdash MN : B[N/x]}&\\
\text{(conv}_\beta\text{)}&\frac{\Gamma \vdash M : A~~~~\Gamma \vdash B : s}{\Gamma \vdash M : B}&A \equiv_\beta B \\
\text{(type/kind} \Pi \text{)}&\frac{\Gamma \vdash A : \star~~~\Gamma, x: A \vdash B : s}{\Gamma \vdash \Pi x : A. B : s}&\\
\text{(var)}&\frac{\Gamma \vdash A : s}{\Gamma, x: A \vdash x : A}&\text{if } x \notin \Gamma
\end{array}
\end{equation}

In 1970, de Bruijn introduced \textsc{Automath} to represent mathematical theorems and proofs. In this system,
a predicate on type $A$ is represented as $P : A \to \star$. For $a \in A$, $P a$ is valid if it is inhabited.
$\forall x\in A. Px$ is translated as $\Pi x: A. Px$ and $A \to B$ is translated as $\Pi x: A. B$.

For example, the following formula is valid: $(\forall x: A. \forall y: A. Pxy) \to (\forall x\in A. Pxx)$ because 
the type $((\Pi x: A. \Pi y: A. Pxy) \to (\Pi x: A. P x x))$ is inhabited, for example by the 
term $\lambda z: (\Pi x: A. \Pi y: A. P xy). \lambda x: A. z x x$.

\subsection{The Cube}
Now, we can consider all three extensions to the ST \lc: terms depending on types, types depending on terms and types
depending on types.

The cube is parametrized by $\mathcal{R}$: which dependencies are allowed. 
$\mathcal{R} \in \mathcal{P}({\star, \square}\times{\star, \square})$

For example, $\lambda P2$ which induces second order predicate logic has $\mathcal{R} = {(\star, \star), (\square, \star), (\star, \square)}$.
The calculus of constructions $\lambda P \omega$ has $\mathcal{R} = {(\star, \star), (\square, \star), (\star, \square), (\square, \square)}$
and corresponds to higher order predicate logic.
$\lambda \omega$ (the not-weak one) has $\mathcal{R} = {(\star, \star), (\square, \star), (\star, \square)}$. 
Actually, what makes \wlo ``weak'' is that it does not have dependent types.

Each of these systems was historically developed separately. They were unified to one cube in the 1980s.

\subsection{Exercises}

\subsubsection{Exercise 1}

\textbf{Prove that $\alpha$-conversion is an equivalence relation.}

\emph{Solution.} We need to show three properties: symmetry, reflexivity, transitivity.
\begin{itemize}
\item \textbf{Symmetry}. Assume that $e_1 \equiv_\alpha e_2$. Then $e_2$ can be reached from $e_1$ by using 
an injective name substitution. Then, $e_1$ can be reached from $e_2$ by using the inverse of that subsitution.
For example, $\lambda x. x \equiv_\alpha \lambda y. y$ using ${x \mapsto y}$, and so $\lambda y. y \equiv_\alpha \lambda x. x$
using ${y \mapsto x}$.
\item \textbf{Reflexivity}. $e \equiv_\alpha e$ using the trivial subsitution ${}$.
\item \textbf{Transitivity}. Assume that $e_1 \equiv_\alpha e_2$ using a name subsitution $f$ and $e_2 \equiv_\alpha e_3$
using a name substitution $g$. Then applying $f$ and then $g$ proves the equivalence $e_1 \equiv_\alpha e_3$.
% Consider $g\bar(x)$ which is equal to $g(x)$ for all inputs for which $g(x)$ is defined,
% and otherwise $g\bar(x)=x$. Then the substitution $g\bar \circ f$ shows that $e_1 \equiv_\alpha e_3$.
% For example, $\lambda x. \lambda y. xy \equiv_\alpha \lambda x. \lambda z. xz$ by $f = {y \mapsto z}$ and
% $\lambda x. \lambda z. xz \equiv_\alpha \lambda p. \lambda k. pk$ by $g = {x \mapsto p, z \mapsto k}$.
% Then, $g\bar = {x \mapsto p, z \mapsto k, y \mapsto y}$
% too much rules-lawyering
\end{itemize}
\qed

\textbf{Prove that every $\lambda$-term has at most one normal form.}

\emph{Solution.} Assume that some $\lambda$-term $e$ has two different normal forms $e_1$ and $e_2$. Then, $e \xreduces{\beta} e_1$ and 
$e \xreduces{\beta} e_2$. According to the Church-Rosser theorem, there is an $e'$ such that $e_1 \xreduces{\beta} e'$ and
$e_2 \xreduces{\beta} e'$. This is in contradiction to $e_1$ and $e_2$ being different normal forms. Notice that this works even if
$e' = e_1$ or $e' = e_2$.
\qed

\textbf{Prove that for every $\lambda$-term $F$, $F (\textbf{Y} F) \equiv_\beta \textbf{Y} F$}

\emph{Solution.} Since $\equiv_\beta$ is symmetric, we'll prove it the other way around.
\[
\begin{array}{rl}
\textbf{Y} F & = (\lambda f. (\lambda x. f (x x)) (\lambda x. f (x x))) F \\
& \equiv_\beta (\lambda x. F (x x)) (\lambda x. F (x x)) \\
& \equiv_\beta F ((\lambda x. F (x x)) (\lambda x. F (x x)))  \\
& \equiv_\beta F (\textbf{Y} F) 
\end{array}
\]
\qed

\textbf{Prove that $((\alpha \to \beta) \to \alpha) \to \alpha$ is not inhabited in $\lambda \to$.}

\emph{Solution.}














